Letter: Performance models for \\ high-resolution ice sheet simulations

### 2022-06-24 paper/perfmod.tex
progress on performance model text itself

Let $s(t,\bx)=H(t,\bx)+b(\bx)$ denote the surface elevation for bed elevation $b(\bx)$.  If the numerical thickness is perturbed then the velocity $\bu$ often responds by growing in the downhill direction $-\gradx s$, a direction often correlated to $-\gradx H$.  In membrane-stress-resolving models this happens through the nonlocal solution of the stress balance; the driving stress is along $\gradx s$.  Equation \eqref{eq:masscontinuity} has velocity $\bu$ which is really a non-local function ``$\bu(H,\grad_x s)$''; a solution of the stress balance equations is required to evaluate this function.

Thus the mass continuity equation is substantially, though not wholely, diffusive.  The diffusive nature dominates in the small-parameter limit which generates the SIA, in which a revised form of the mass continuity equation holds:
\begin{equation}
\frac{\partial H}{\partial t} = \Divx \left(D \gradx s \right) + a  \label{eq:siamasscontinuity}
\end{equation}
where
\begin{equation}
D = \Gamma H^{n+2} |\gradx s|^{n-1} \label{eq:siadiffusivity}
\end{equation}
is the nonlinear diffusivity.  Here $\Gamma = 2 A (\rho g)^n/(n+2)$ is a positive constant in the isothermal case, with $A$ the ice softness, $\rho$ the ice density, $g$ gravity, and $n$ the Glen exponent in the flow law \citep{GreveBlatter2009}.

Unlike equation \eqref{eq:masscontinuity}, equation \eqref{eq:siamasscontinuity} does not hold for all models, but the essentially-geometric diffusivity $D$ can be computed from equation \eqref{eq:siadiffusivity}, regardless.  In any numerical model of grounded ice sheets, large values of $D$ will coincide with locations of unstable mode growth if explicit time-steps are chosen too large.

FIXME no characteristics

FIXME SITUATION COMPLICATED AND PRONE TO MISUNDERSTANDING; NEED PERFORMANCE MODEL TO UNDERSTAND TRADE-OFFS; We expect that time-stepping glacier and ice-sheet numerical models will become faster, and thus more useful tools for glaciologists, when multigrid is applied to the solution of the problem(s) solved at each time-step.  This hope can be understood and assessed by a \emph{performance model} for the scheme, one which combines computational work estimates for subproblems into an asymptotic, in the limit of highly-refined meshes, description of of the work required for a given simulation task.

@@

Consider numerical simulations based on $n$ mesh nodes in the map-plane, i.e.~$n$ total horizontal grid points.  (Table \ref{tab:notation} lists all of our notation.)  Such a simulation uses $O(n)$ surface elevation variables, the \emph{degrees of freedom} in our analysis.  It requires $O(n)$ memory to store the instantaneous model state, including geometry, velocity, and thermodynamical variables, if we assume that the vertical grid/mesh has an \emph{a priori} bounded vertical resolution, a common \citep[for example]{Brinkerhoffetal2017,Hoffmanetal2018}, though not universal \citep{IsaacStadlerGhattas2015}, property of such models.

@@

The glacier simulation covers a model domain of horizontal span $L>0$.  For concreteness, in 1D the domain is an interval $[0,L]$, while in 2D it is $[0,L]^2$, or a subset thereof.  Let $\Delta x$ be the characteristic width (diameter) of map-plane mesh cells, thus $\Delta x = O(L n^{-1})$ in 1D and $\Delta x = O(L n^{-1/2})$ in 2D.

Ice sheet models involve climate interaction, especially SMB, and the interaction is dominated by the annual cycle.  Let $q$ be the number of time steps per year needed to incorporate this climate interaction.  Typical values $q=1,12,365 \,\text{a}^{-1}$ correspond to yearly, monthly, and daily climatic interaction frequency.

Almost all current-technology glacier and ice-sheet models use explicit time-stepping which is only conditionally stable \citep{LeVeque2007}.  We will assume that for the resolutions used in scientific applications, maintenance of the stability of explicit time-stepping requires substantially more than $Q$ time steps per year.

For an explicit SIA model the well-known stability restriction is $\Delta t < O(D^{-1} \Delta x^2)$ where $D$ is a representative effective diffusivity in $\text{km}^2\,\text{a}^{-1}$ \citep{Bueleretal2005,HindmarshPayne1996}.
For Stokes dynamics the question of explicit time-stepping stability is largely unexplored.  We propose a \emph{pessimistic} explicit time-step stability paradigm, namely the above SIA restriction $\Delta t < O(D^{-1} \Delta x^2)$.  This is motivated by the small geometric aspect ratio analysis of the Stokes model leading to the SIA model \citep{GreveBlatter2009}.  Alternatively one might subscribe to an \emph{optimistic} paradigm supposing that stability is controlled by the advective restriction $\Delta t < O(U^{-1} \Delta x)$ for some typical horizontal velocity scale $U$ in $\text{km}\,\text{a}^{-1}$.

So-called ``higher-order'' and ``hybrid'' are perhaps better-studied than numerical Stokes models, especially over a large range of horizontal resolutions.  Certain ``hybrid'' schemes simply use the SIA-type stability restriction \citep{Winkelmannetal2011}.  The optimistic concept above is partially supported by a recent analysis of a certain higher-order scheme (``DIVA'') for which $\Delta t < \min\left\{C,O(U^{-1} \Delta x)\right\}$ \citep[equations (52) and (56)]{Robinsonetal2022}, wherein $C>0$ is independent of $\Delta x$.  However, practical Greenland DIVA simulations in the same work actually suggest $\Delta t = O(\Delta x^{1.6})$ instead \citep[Figure 3(a)]{Robinsonetal2022}.

For fixed geometry, as in each time step of an explicit simulation, the momentum balance equations generally require nontrivial solutions as PDEs.  For the Stokes model we suppose one such solution requires $O(n^{1+\alpha})$ work.  Note that such a scheme will determine the surface velocity appearing in the SKE.  However, the value of $\alpha \ge 0$ depends strongly on the numerical velocity/pressure solver implementation, that is, the numerics by which the Stokes momentum balance is actually solved.  Naive solution might involve $\alpha=2$, for example if dense linear algebra is involved, but application of geometric multigrid methods by \cite{IsaacStadlerGhattas2015}, with a solver specifically adapted to the geometry and flow law applicable to ice sheets, suggest that $\alpha \approx 0$ is potentially attainable.  (Of course, the constant $C$ in $(\text{work}) \le C n^{1+\alpha}$ is large.)  Geometric multigrid methods have also been applied to the higher-order (hydrostatic) momentum balance equations \citep{BrownSmithAhmadia2013}, a shallow approximation which resolves both vertical shear stresses and membrane stresses, and in this case a velocity solution work estimate $O(n^{1+\alpha})$ with $\alpha \approx 0$ is a reasonable aspiration.

The SIA model velocity computation is a trivialization of the Stokes problem.  Each velocity unknown is computed in $O(1)$ work, thus the velocity solution is attained (from fixed geometry) in $O(n)$ work.

FIXME \citep{LofgrenAhlkronaHelanow2021} suggest how to reduce the constant in explicit methods for Stokes geometry updates

On the other hand, an unconditionally-stable implicit geometric-update scheme, e.g.~as demonstrated in \citep{Bueler2016} for the SIA, will take at most $Q$ steps per model year in order to follow the climate forcing.  For such an implicit scheme, our stability model is that there is no restriction in the fine-resolution limit ($\Delta x \to 0$).  For example, the steady-state SIA solver in \cite{Bueler2016} uses $\Delta t=+\infty$, with $\Delta t = 100 \,\text{a}$ as a recovery strategy in the case of solver non-convergence.

@@

\newcommand{\oo}[1]{{\LARGE \strut} \displaystyle O\left(#1\right)}
\setlength{\tabcolsep}{5pt}
\renewcommand{\arraystretch}{1.5}
\begin{table*}[ht]
\begin{tabular}{lll}
\emph{time-stepping} & \emph{dynamics} & \emph{work (flops per model year)} \\ \hline
explicit & SIA    & 1D:\quad $\oo{\frac{D}{L^2} n^3}$ \\
         &        & 2D:\quad $\oo{\frac{D}{L^2} n^2}$ \\
explicit & Stokes & 1D:\quad $\oo{\frac{U}{L} n^{2+\alpha}}$ \quad $\left[\oo{\frac{D}{L^2} n^{3+\alpha}} \text{if pessimistic}\right]$ \\
         &        & 2D:\quad $\oo{\frac{U}{L} n^{1.5+\alpha}}$ \quad $\left[\oo{\frac{D}{L^2} n^{2+\alpha}} \text{if pessimistic}\right]$ \\
implicit & SIA    & $\oo{q\, n^{1+\beta}}$ \\
implicit & Stokes & $\oo{q\, n^{1+\gamma}}$
\end{tabular}
\caption{Asymptotic scaling of computational work for time-stepping numerical ice sheet simulations as the degrees of freedom $n$ goes to infinity (high horizontal resolution).  Note 1D and 2D refer to flow-line and map-plane simulations, respectively.  See Table \ref{tab:notation} for remaining notation.}
\label{tab:performancemodel}
\end{table*}

>>>>

Let $s(t,\bx)=H(t,\bx)+b(\bx)$ denote the surface elevation for bed elevation $b(\bx)$.  If the numerical thickness is perturbed then the velocity $\bu$ often responds by growing in the downhill direction $-\gradx s$, a direction often correlated to $-\gradx H$.  In membrane-stress-resolving models this happens through the nonlocal solution of the stress balance; the driving stress is along $\gradx s$.  Equation \eqref{eq:masscontinuity} has velocity $\bu$ which is really a non-local function ``$\bu(H,\grad_x s)$''; a solution of the stress balance equations is required to evaluate this function.  Equivalently, ice sheet flow has no characteristic curves because the velocity $\bu$ actually depends indirectly on the gradient of thickness.

Thus the mass continuity equation is substantially, though not wholely, diffusive.  The diffusive nature dominates in the small-parameter limit which generates the SIA, in which a revised mass continuity equation holds:
\begin{equation}
\frac{\partial H}{\partial t} = \Divx \left(d\, \gradx s \right) + a, \,  \label{eq:siamasscontinuity}
\end{equation}
where $d = C H^{n+2} |\gradx s|^{n-1}$ is the nonlinear diffusivity, $C = 2 A (\rho g)^n/(n+2)$ is a positive constant, $A$ is the (isothermal) ice softness, $\rho$ is the ice density, $g$ is gravity, and $n$ is the Glen exponent in the flow law \citep{GreveBlatter2009}.

Unlike equation \eqref{eq:masscontinuity}, equation \eqref{eq:siamasscontinuity} does not hold for all models.  Regardless, the same diffusivity $d$, an essentially geometric quantity, can be computed.  For grounded ice sheets large values of $d$ will indicate locations of unstable mode growth if explicit time-steps are chosen too large.

The present-day understanding of ice sheet model time stepping is complicated.  While the author expects that time-stepping ice sheet models will eventually become faster when implicit time-stepping and advanced solver techniques like multigrid \citep{Briggsetal2000} are applied, this belief should be inspected quantitatively to the extent possible.  A simplified performance model can expose the most important considerations and trade-offs.

@@

We analyze simulation performance using $n$, the number of mesh/grid nodes in the horizontal map-plane.  (Table \ref{tab:notation} lists our notation.)  A numerical ice sheet model uses $n$ ice thickness or surface elevation variables, the number of \emph{degrees of freedom}.  It requires $O(n)$ memory to store the instantaneous model state, including thermodynamical variables for example, if we assume that the vertical mesh/grid has an \emph{a priori} bounded resolution.\footnote{This bounded-vertical-nodes assumption reflects common usage \citep[for example]{Aschwandenetal2019,Brinkerhoffetal2017,Hoffmanetal2018}, but it may hold only in an approximate sense \citep{IsaacStadlerGhattas2015}.}  The amount of fast memory needed by a running ice sheet simulation is also $O(n)$ if prior states are discarded or transferred to storage.

@@

Ice sheet models also have $O(n)$ velocity variables.  However, these are not state variables because a very-viscous stress balance can recompute velocity as a function of (true) state variables.

The model domain is of (horizontal) width $L>0$.  If $\Delta x$ is a representative diameter of the map-plane mesh cells then $\Delta x = O(L n^{-1/2})$; there are $O(\sqrt{n})$ mesh cells in each horizontal dimension.  In flow-line models, not our primary motivation but included for completeness, $\Delta x = O(L n^{-1})$.

Ice sheet models involve climate interactions, especially via surface mass balance, which are dominated by an annual cycle and longer time-scale changes.  Let $q$ be the number of ice-dynamical time steps per year needed to capture this climate interaction; typical values $q=0.1 \,\text{a}^{-1}, 1 \,\text{a}^{-1}, 12 \,\text{a}^{-1}$ correspond to decadal, yearly, and monthly flow/climate interaction frequency, respectively.  Energy balance and degree-day schemes for computing surface mass balance generally have much shorter time scales, but here $q$ only describes the frequency on which ice geometry is updated from an ice velocity solution.

Current-technology ice sheet models mostly use explicit time-stepping which is only conditionally stable \citep{LeVeque2007}.  For the spatial resolutions used in present-day scientific applications, we observe that maintenance of explicit time-stepping stability requires time steps substantially shorter than $1/q$ model years.  Unconditionally-stable implicit schemes also have a maximum time step restriction, namely $1/q$ itself, reflecting the simulation purpose and not the maintenance of stability.

For an explicit SIA model the well-known stability restriction is $\Delta t < O(D^{-1} \Delta x^2)$ \citep{Bueleretal2005,HindmarshPayne1996} where $D$ is a representative diffusivity value from the diffusivity formula in equation \eqref{eq:siamasscontinuity}.

For Stokes dynamics the stability of explicit time-stepping is largely unexplored in any precise sense \citep[compare][]{Chengetal2017}.  We propose that an advective restriction $\Delta t < O(U^{-1} \Delta x)$, for some representative horizontal velocity scale $U$, computed from simulated ice velocities, represents the \emph{optimistic} paradigm.  The corresponding \emph{pessimistic} paradigm enforces $\Delta t < O(D^{-1} \Delta x^2)$, using diagnostic diffusivity values computed in the SIA manner.

Explicit time-stepping with hybrid and higher-order schemes is better-studied than for Stokes dynamics, especially over horizontal resolutions relevant to whole ice sheets.  Some hybrid schemes apply the pessimistic time step as an adaptive restriction \citep{Winkelmannetal2011}.  The optimistic time step is supported for a certain higher-order scheme \citep[``DIVA'';][equation (52)]{Robinsonetal2022}, but practical Greenland simulations in the same work actually suggest $\Delta t = O(\Delta x^{1.6})$ (Figure 3(a)).

Regardless of time-step stability, solution of the momentum balance equations requires effort.  In general this requires a nontrivial solution of spatial partial differential equations.  For the Stokes model, in particular, we suppose one such solution requires $O(n^{1+\alpha})$ work, namely floating point operations in our simple performance model, where $\alpha\ge 0$ depends on the design of the stress balance solver.  Naive solution using direct linear algebra might involve $\alpha=3$, or $\alpha \ge 2$ when exploiting sparsity.  However, application of a multigrid method, such as the \cite{IsaacStadlerGhattas2015} solver specifically adapted to the geometry and flow law applicable to ice sheets, suggest that $\alpha$ can potentially be close to zero, with the important caveat that the constant implied in ``$O(n^{1+\alpha})$'' is very large.  Multigrid methods have also been applied to higher-order stress balance equations \citep{BrownSmithAhmadia2013}; again $\alpha \approx 0$ is a credible aspiration.  On the other hand, the SIA velocity computation is a trivialization of the Stokes problem; here each velocity is computed in $O(1)$ work, and the velocity solution in $O(n)$ work.

The stress-balance scheme determines the velocity appearing in equation \eqref{eq:masscontinuity}.  From the velocity field, regardless of the stress balance source, an explicit time-stepping scheme computes the updated ice thickness using $O(n)$ work.

On the other hand, an unconditionally-stable implicit geometric-update scheme in the SIA, as demonstrated by \cite{Bueler2016}, FIXME $O(n^{1+\beta})$ work; for Stokes this is a different task of solving coupled equations and we merely write $O(n^{1+\gamma})$ for now

FIXME $\Delta t^{-1}$ time steps per model year (if $\Delta t$ is in model years)

For such an implicit scheme, our stability model is that there is no restriction in the fine-resolution limit ($\Delta x \to 0$).  For example, the steady-state SIA solver in \cite{Bueler2016} uses $\Delta t=+\infty$, with $\Delta t = 100 \,\text{a}$ as a recovery strategy in the case of solver non-convergence.

@@

\newcommand{\oo}[1]{{\LARGE \strut} \displaystyle O\left(#1\right)}
\setlength{\tabcolsep}{5pt}
\renewcommand{\arraystretch}{1.5}
\begin{table*}[ht]
\begin{tabular}{llll}
\emph{time-stepping} & \emph{dynamics} & \emph{flops per model year \, $\left[\text{1D}\right]$} & \emph{pessimistic \, $\left[\text{1D}\right]$} \\ \hline
explicit & SIA    & $\oo{\frac{D}{L^2} n^2}$ \, $\left[n^3\right]$ \\
explicit & Stokes & $\oo{\frac{U}{L} n^{1.5+\alpha}}$ \, $\left[n^{2+\alpha}\right]$ & $\oo{\frac{D}{L^2} n^{2+\alpha}}$ \, $\left[n^{3+\alpha}\right]$ \\
implicit & SIA    & $\oo{q\, n^{1+\beta}}$ \\
implicit & Stokes & $\oo{q\, n^{1+\gamma}}$
\end{tabular}
\caption{Asymptotic scaling of computational work, floating point operations per model year, for time-stepping numerical ice sheet simulations, as the degrees of freedom $n$ goes to infinity.  Note 1D refers to flow-line geometry; corrected power on $n$ for 1D is in square brackets.  See Table \ref{tab:notation} for remaining notation.}
\label{tab:performancemodel}
\end{table*}
